Article Title

### 2013-08-19 sections/Estimators.tex
changed the format to PLOS computational biology

\subsection{Estimator A: Shrinkage toward diagonal}
The most popular covariance regularization schemes use diagonal target estimates and linear shrinkage toward the target.  For example, the target could be the identity matrix:  
\begin{equation}
\hat T = \hat v I
\end{equation}
where $\hat v = \frac 1 p \sum\limits_{i=1}^p(\hat\Sigma_0)_{ii}$ is the mean sample variance.

@@

The hyperparameters $\hat\eta$ and $\hat\lambda$ must be estimated from the data.  Under the MSE loss function $\mathcal L_e$ (\autoref{eq:MSE}), the optimal values of $\hat\eta$ and $\hat\lambda$ can be estimated analytically \citep{Ledoit:2004,Schafer:2005,Schaefer:2010}. However, these estimates are no longer optimal under the Guassian loss $\mathcal L_g$. \TODO{I have tested this and there was a substantial difference in both synthetic and empirical data}  When an analytical solution for optimal hyperparameters is not available, the optimal values can be found by cross validation, entirely within the the training sample. \TODO{explain nested cross-validation in more detail?}

@@

\subsection{Estimator B: Shrinkage toward a factor model}
In estimator B, the target is a factor model, composed as the sum of a low-rank component $\hat L_{\hat d}\hat L_{\hat d}^\T$ and a diagonal matrix $\hat \Psi$:
\begin{equation}
\hat T_{\hat d} = \hat L_{\hat d} \hat L_{\hat d}^\T + \hat\Psi
\end{equation}
Here $\hat L_{\hat d}$ is an $p\times\hat d$ matrix and $\hat\Psi$ is diagonal.

If $F$ were assume to be multivariate normal \TODO{or other distributions that are defined by linear dependencies}, the factor model is corresponds to the graphical model (\autoref{fig:02}B) where the activity of all neurons are dependent only $\hat d$ latent units. This suggest a mechanistic interpretation in which the interactions between neurons are insignificant compared to the influence of inputs outside the recorded circuit.

@@

\TODO{Check \citep{Ledoit:2003,Fan:2011,Fan:2006}}

\subsection{Estimator C: Sparse inverse}
Assuming that $x$ is distributed normally, the inverse of the covariance matrix or \emph{precision matrix} $K=\Sigma^{-1}$ has special significance: zeros in the precision matrix indicate conditional independence between the corresponding pairs. To see this, let $x=(x_1,\ldots,x_p)^\T \sim g\left(x\cond K\right)$
\begin{equation}
g\left(x\cond K\right) = \frac 1 {Z(K)} \exp\left(-\frac 1 2 \sum\limits_{i=1}^p\sum\limits_{j=1}^p  K_{ij} x_i x_j\right)
\end{equation} 
If $K_{12}\equiv K_{21} = 0$, then the joint distribution of $x_1$ and $x_2$, conditioned on $x_3=a_3,\ldots,x_p=a_p$ can be decomposed as product of independent distributions of $x_1$ and $x_2$: 
\begin{equation}
\begin{split}
g\left( x_1, x_2 \cond K, x_3=a_3,\ldots,x_p=a_p\right) &=
\frac 1 {Z(K)} \exp\left(-\frac 1 2 \sum\limits_{i=3}^p\sum\limits_{j=3}^p  K_{ij} a_i a_j \right)\times
\\
 &  \exp\left( -\frac 1 2\left( K_{11} x_1^2 +  x_1\sum\limits_{i=3}^p K_{1i}a_i \right)  \right)
\\
 &  \exp\left( -\frac 1 2\left( K_{22} x_2^2 +  x_2\sum\limits_{i=3}^p K_{2i}a_i \right)  \right)
\end{split}
\end{equation}

\citep{Dempster:1972,Meinshausen:2006,Friedman:2008}

\subsection{Estimator D: Sparse inverse with latent units}
\citep{Ma:2013}

>>>>

\subsection*{Estimator A: Shrinkage toward diagonal}
The most popular covariance regularization schemes use diagonal target estimates and linear shrinkage toward the target.  For example, the target could be the identity matrix:  
\begin{equation}
\hat T = \hat v I
\end{equation}
where $\hat v = \frac 1 p \sum\limits_{i=1}^p(\hat\Sigma_0)_{ii}$ is the mean sample variance.

@@

The hyperparameters $\hat\eta$ and $\hat\lambda$ must be estimated from the data.  Under the MSE loss function $\mathcal L_e$ (\ref{eq:MSE}), the optimal values of $\hat\eta$ and $\hat\lambda$ can be estimated analytically \cite{Ledoit:2004,Schafer:2005,Schaefer:2010}. However, these estimates are no longer optimal under the Guassian loss $\mathcal L_g$. \TODO{I have tested this and there was a substantial difference in both synthetic and empirical data}  When an analytical solution for optimal hyperparameters is not available, the optimal values can be found by cross validation, entirely within the the training sample. \TODO{explain nested cross-validation in more detail?}

@@

\subsection*{Estimator B: Shrinkage toward a factor model}
In estimator B, the target is a factor model, composed as the sum of a low-rank component $\hat L_{\hat d}\hat L_{\hat d}^\T$ and a diagonal matrix $\hat \Psi$:
\begin{equation}
\hat T_{\hat d} = \hat L_{\hat d} \hat L_{\hat d}^\T + \hat\Psi
\end{equation}
Here $\hat L_{\hat d}$ is an $p\times\hat d$ matrix and $\hat\Psi$ is diagonal.

If $F$ were assume to be multivariate normal \TODO{or other distributions that are defined by linear dependencies}, the factor model is corresponds to the graphical model (\ref{fig:02}B) where the activity of all neurons are dependent only $\hat d$ latent units. This suggest a mechanistic interpretation in which the interactions between neurons are insignificant compared to the influence of inputs outside the recorded circuit.

@@

\TODO{Check \cite{Ledoit:2003,Fan:2011,Fan:2006}}

\subsection*{Estimator C: Sparse inverse}
Assuming that $x$ is distributed normally, the inverse of the covariance matrix or \emph{precision matrix} $K=\Sigma^{-1}$ has special significance: zeros in the precision matrix indicate conditional independence between the corresponding pairs. To see this, let $x=(x_1,\ldots,x_p)^\T \sim g\left(x\cond K\right)$
\begin{equation}
g\left(x\cond K\right) = \frac 1 {Z(K)} \exp\left(-\frac 1 2 \sum\limits_{i=1}^p\sum\limits_{j=1}^p  K_{ij} x_i x_j\right)
\end{equation} 
If $K_{12}\equiv K_{21} = 0$, then the joint distribution of $x_1$ and $x_2$, conditioned on $x_3=a_3,\ldots,x_p=a_p$ can be decomposed as product of independent distributions of $x_1$ and $x_2$: 
\begin{equation}
\begin{split}
g\left( x_1, x_2 \cond K, x_3=a_3,\ldots,x_p=a_p\right) &=
\frac 1 {Z(K)} \exp\left(-\frac 1 2 \sum\limits_{i=3}^p\sum\limits_{j=3}^p  K_{ij} a_i a_j \right)\times
\\
 &  \exp\left( -\frac 1 2\left( K_{11} x_1^2 +  x_1\sum\limits_{i=3}^p K_{1i}a_i \right)  \right)
\\
 &  \exp\left( -\frac 1 2\left( K_{22} x_2^2 +  x_2\sum\limits_{i=3}^p K_{2i}a_i \right)  \right)
\end{split}
\end{equation}

\cite{Dempster:1972,Meinshausen:2006,Friedman:2008}

\subsection*{Estimator D: Sparse inverse with latent units}
\cite{Ma:2013}

### 2013-08-19 sections/Introduction.tex
changed the format to PLOS computational biology

Pearson correlations between the spiking activity of pairs of neurons, or simply \emph{neural correlations}, are among the most familiar descriptive statistics of neural population activity \citep{Averbeck:2006,Zohary:1994,Kohn:2005,Bair:2001,Renart:2010}.  For example, \emph{noise correlations}, \emph{i.e.}~the correlations of stimulus response variability between pairs of neurons, have been of particular interest.  Noise correlations can have profound theoretical implications for stimulus coding \citep{Zohary:1994,Abbott:1999,Averbeck:2006,Berens:2011,Ecker:2011}, and have been interpreted to indicate detailed and specific functional organization. Such interpretation is supported by a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distance separating the neurons \citep{Smith:2008,Denman:2013}, their synaptic connectivity and stimulus tuning similarity \citep{Kohn:2005,Ko:2011}, cortical layer specificity \citep{Hansen:2012,Smith:2013}, cell-type specificity (?), progressive changes in development and in learning \citep{Golshani:2009}, changes due to sensory stimulation and global brain states \citep{Goard:2009,Kohn:2009,Ecker:2010,Renart:2010}, and others.

However, neural correlations do not come with ready or unambiguous mechanistic interpretations.  Theoretical and simulation studies have shown that neural correlations may arise at various temporal scales from combinations of multiple underlying mechanisms.  These include direct synaptic interactions, common or correlated inputs, shared sensory noise, chains of multiple synaptic connections, oscillations, top-down modulation, and background network activity \citep{Perkel:1967b,Shadlen:1998,Salinas:2001,Ostojic:2009,Rosenbaum:2011}.

Early studies of neural correlations were based on measurements from isolated pairs of neurons and their impact on coding was extrapolated to entire populations \citep{Shadlen:1998,Zohary:1994}.  Multineuronal recordings allow estimation of covariance matrices of large populations of neurons.  Such estimates provide more information than the equivalent number of pairwise correlations assessed in isolation. Indeed, the correlation matrix is greater than the sum of its parts: it can be transformed into other representations that accentuate different aspects of the correlation structure and may suggest different mechanistic interpretations. For example, the eigenvalue decomposition of the covariance matrix expresses shared correlated activity components across the population; common fluctuations of population activity may be accurately represented by just a few principal components but will affect all correlation coefficients. In contrast, the off-diagonal elements of the inverse of the correlation matrix constitute scaled partial correlations between neuron pairs, which reflect their specific linear dependencies, after accounting for the activity of all the other recorded cells; a strong interaction between a pair of neurons may be expressed by a single partial correlation but its effects may propagate to multiple correlations and eigenvalues.   The inverse of the covariance matrix plays an important role in decoding schemes such as linear discriminant analysis, for example.  The mutliple  representations of the covariance matrix with their alternative interpretations add both complications and opportunities into the search for fundamental regularities in neural population activity.

With large numbers of recorded cells, the usual estimations of the covariance matrix from finite recordings become ill-conditioned. Numerical instabilities arise because, as the amount of recorded data increases only linearly, the number of free coefficients to be estimated in any parameterization of the covariance matrix increases quadratically. 
For example, the large eigenvalues of large sample covariance matrices are biased upward while small eigenvalues are biased downward \citep{Hayes:1981}. Similarly, the coefficients of the inverse covariance matrix require larger sample sizes to be estimated accurately from high-dimensional data than low-dimensional data.

With this study, we pursue two related goals: 
\begin{enumerate}[\qquad A.\;\;]
	\item Devise more efficient estimators of neural covariance matrices in recordings from large populations of neurons.
	\item Aid interpretation of population activity  by isolating the essential structure of the neural covariance matrix.
\end{enumerate}

To select the best covariance matrix estimate for a specific neural circuit, we evaluate the performance of four estimators using different regularization schemes, each motivated by  a different hypothesis for the origin of neural correlations. \emph{Regularization} is the deliberate biasing the solution toward a simplified, low-dimensional (`sparse') \emph{target estimate} \citep{Schafer:2005,Bickel:2006}.
Regularized estimators allow reaching a favorable tradeoff between bias and variance in order to minimize the total error.
Strikingly, some improvement can be produced even when the target estimate is chosen arbitrarily, which is sometimes described as ``Stein's phenomenon'' after its discoverer \cite{Stein:1956}.   However, when a sparse target estimate does happen to match the dominant features of the true covariance matrix, it will induce relatively little bias while retaining low variance.  For example, some covariance estimators have been motivated by a specific model of the underlying process such as single-factor models of the stock market, for example \citep{Ledoit:2003}. 
Since the true covariance structure is not known in practice, the performance of estimators is evaluated by how well their performance generalized to new data, outside the training sample by cross validation, for example.

>>>>

Pearson correlations between the spiking activity of pairs of neurons, or simply \emph{neural correlations}, are among the most familiar descriptive statistics of neural population activity \cite{Averbeck:2006,Zohary:1994,Kohn:2005,Bair:2001,Renart:2010}.  For example, \emph{noise correlations}, \emph{i.e.}~the correlations of stimulus response variability between pairs of neurons, have been of particular interest.  Noise correlations can have profound theoretical implications for stimulus coding \cite{Zohary:1994,Abbott:1999,Averbeck:2006,Berens:2011,Ecker:2011}, and have been interpreted to indicate detailed and specific functional organization. Such interpretation is supported by a series of discoveries of nontrivial relationships between neural correlations and other aspects of circuit organization such as the physical distance separating the neurons \cite{Smith:2008,Denman:2013}, their synaptic connectivity and stimulus tuning similarity \cite{Kohn:2005,Ko:2011}, cortical layer specificity \cite{Hansen:2012,Smith:2013}, cell-type specificity (?), progressive changes in development and in learning \cite{Golshani:2009}, changes due to sensory stimulation and global brain states \cite{Goard:2009,Kohn:2009,Ecker:2010,Renart:2010}, and others.

However, neural correlations do not come with ready or unambiguous mechanistic interpretations.  Theoretical and simulation studies have shown that neural correlations may arise at various temporal scales from combinations of multiple underlying mechanisms.  These include direct synaptic interactions, common or correlated inputs, shared sensory noise, chains of multiple synaptic connections, oscillations, top-down modulation, and background network activity \cite{Perkel:1967b,Shadlen:1998,Salinas:2001,Ostojic:2009,Rosenbaum:2011}.

Early studies of neural correlations were based on measurements from isolated pairs of neurons and their impact on coding was extrapolated to entire populations \cite{Shadlen:1998,Zohary:1994}.  Multineuronal recordings allow estimation of covariance matrices of large populations of neurons.  Such estimates provide more information than the equivalent number of pairwise correlations assessed in isolation. Indeed, the correlation matrix is greater than the sum of its parts: it can be transformed into other representations that accentuate different aspects of the correlation structure and may suggest different mechanistic interpretations. For example, the eigenvalue decomposition of the covariance matrix expresses shared correlated activity components across the population; common fluctuations of population activity may be accurately represented by just a few principal components but will affect all correlation coefficients. In contrast, the off-diagonal elements of the inverse of the correlation matrix constitute scaled partial correlations between neuron pairs, which reflect their specific linear dependencies, after accounting for the activity of all the other recorded cells; a strong interaction between a pair of neurons may be expressed by a single partial correlation but its effects may propagate to multiple correlations and eigenvalues.   The inverse of the covariance matrix plays an important role in decoding schemes such as linear discriminant analysis, for example.  The mutliple  representations of the covariance matrix with their alternative interpretations add both complications and opportunities into the search for fundamental regularities in neural population activity.

With large numbers of recorded cells, the usual estimations of the covariance matrix from finite recordings become ill-conditioned. Numerical instabilities arise because, as the amount of recorded data increases only linearly, the number of free coefficients to be estimated in any parameterization of the covariance matrix increases quadratically. 
For example, the large eigenvalues of large sample covariance matrices are biased upward while small eigenvalues are biased downward \cite{Hayes:1981}. Similarly, the coefficients of the inverse covariance matrix require larger sample sizes to be estimated accurately from high-dimensional data than low-dimensional data.

With this study, we pursue two related goals: 
\begin{enumerate}
	\item Devise more efficient estimators of neural covariance matrices in recordings from large populations of neurons.
	\item Aid interpretation of population activity  by isolating the essential structure of the neural covariance matrix.
\end{enumerate}

To select the best covariance matrix estimate for a specific neural circuit, we evaluate the performance of four estimators using different regularization schemes, each motivated by  a different hypothesis for the origin of neural correlations. \emph{Regularization} is the deliberate biasing the solution toward a simplified, low-dimensional (`sparse') \emph{target estimate} \cite{Schafer:2005,Bickel:2006}.
Regularized estimators allow reaching a favorable tradeoff between bias and variance in order to minimize the total error.
Strikingly, some improvement can be produced even when the target estimate is chosen arbitrarily, which is sometimes described as ``Stein's phenomenon'' after its discoverer \cite{Stein:1956}.   However, when a sparse target estimate does happen to match the dominant features of the true covariance matrix, it will induce relatively little bias while retaining low variance.  For example, some covariance estimators have been motivated by a specific model of the underlying process such as single-factor models of the stock market, for example \cite{Ledoit:2003}. 
Since the true covariance structure is not known in practice, the performance of estimators is evaluated by how well their performance generalized to new data, outside the training sample by cross validation, for example.

### 2013-08-19 sections/Theory.tex
changed the format to PLOS computational biology

\subsection{Basic concepts}
Let $x(t) \in \mathcal X, t=1\ldots,n$ denote a sample of consecutive observations of population activity of $p$ neurons in time bins $t$.  
The activity of each neuron is denoted by the real-valued firing rate, thus  $\mathcal X = \mathbb R^{p\times 1}$.  
For example, in our applicaiton, the firing rate is estimated from somatic calcium fluorescence signals with the mean stimulus response subtracted, making it possible for elements of $x(t)$ to take on negative and non-integer values.

@@

But the usual estimator of the true covariance matrix $\Sigma$ is  the sample covariance matrix:
\begin{equation}
\hat\Sigma_0 =  \frac 1 {n-c} \sum\limits_{t=1}^n (x(t) -\hat\mu) (x(t) - \hat\mu)^\T
\quad\mbox{where}\;
\hat\mu = \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $c$ is the sum of temporal correlations \TODO{there must be a better description of this}. For independent observations, $c=1$;  $c>1$ when nearby samples are correlated; the value of $c$ can be estimated from the data.
\TODO{Under the Gaussian loss, $\mathcal L_g$ (\autoref{eq:GaussLoss}) the scaling is irrelevant, so we will not spend any time talking about the scaling.  But it would be important if we used $\mathcal L_e$ (\autoref{eq:MSE}).}

@@

Nonlinear functions of an unbiased estimate can be biased: $\mathbb E\left[\hat\Sigma_0\right]=\Sigma \centernot\implies  E\left[\varphi(\hat\Sigma_0)\right]=\varphi(\Sigma)$.  For example, if $u_{\max}(S)$ is the largest eigevalue of square matrix $S$, then $\mathbb E\left[u_{\max}(\hat\Sigma_0)\right] > u_{\max}(\Sigma)$. Some covariance matrix estimators are designed to correct for the eigenspectrum bias at the cost of adding bias to covariance coefficients \citep{Ledoit:2004}.

\subsection{Loss and risk}
The optimization of a covariance matrix estimate $\hat\Sigma$ is performed with respect to a \emph{loss function} $\mathcal L(\hat\Sigma,\Sigma)$, which expresses the discrepancy between $\hat\Sigma$ and $\Sigma$ and attains its minimum when $\hat\Sigma=\Sigma$.  
Then \emph{excess loss}  
\begin{equation}
\ell(\hat\Sigma,\Sigma) = \mathcal L(\hat\Sigma,\Sigma)-\mathcal L(\Sigma,\Sigma)
\end{equation}
assumes zero at its minimum.

@@

The Gaussian loss function $\mathcal L_g$  arises from the theory of multivariate normal distributions. When observations are identically and independently distributed according to a multivariate normal distribution with zero means, the log likelihood of the covariance matrix $\Sigma$ with $\hat\Sigma = \frac 1 n \sum\limits_{t=1}^n x(t) x(t)^\T$ is  
\begin{equation}
L\left(\Sigma \mid \hat\Sigma\right) = -\frac n 2 \ln(2\pi) - \frac n 2 \ln \det \Sigma - \frac n 2 \Tr(\Sigma^{-1} \hat \Sigma)
\end{equation}
Then $\mathcal L_g$ is constructed by rescaling $L\left(\Sigma \mid \hat\Sigma\right)$ and dropping the constant term:
\begin{equation}\label{eq:GaussLoss}
\mathcal L_g(\hat\Sigma,\Sigma) 
=  -\frac 2 {pn} L\left(\Sigma \mid \hat\Sigma \right) - \frac 1 p \ln(2\pi) 
\equiv  \frac 1 p\left(\ln \det \hat \Sigma + \Tr(\hat \Sigma^{-1}) \right) 
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat\Sigma,\Sigma) = \mathcal L_g(\hat\Sigma,\Sigma) - \mathcal L_g(\Sigma,\Sigma)  
= \frac 1 p \left(-\ln \det (\hat \Sigma^{-1} \Sigma) + \Tr(\hat \Sigma^{-1}\Sigma)\right) - 1
\end{equation}
is known as \emph{entropy loss} \citep{James:1961}. \TODO{The order of the operands is reversed. Check that it's okay.}

@@

\subsection{Cross validation}
In practice, the true covariance matrix $\Sigma$ is not accessible and the estimator risk $r$ must be estimated from data. This requires a separate \emph{validation} empirical distribution $\hat F_m^\prime$ of $m$ observations sampled from $F$ independently of the \emph{training} distribution $\hat F_n$. 
Then let  $\hat \Sigma_0^\prime$ be the sample covariance obtained from $\hat F_m^\prime$.

@@

\subsection{Bias/variance decomposition}
The estimator risk (\autoref{eq:risk}) can be decomposed into \emph{approximation error} or \emph{bias}   
\begin{equation}
b^2 = \ell \left( \mathbb E[\hat \Sigma],\Sigma\right)
\end{equation}
and \emph{estimation error} or \emph{variance}
\begin{equation}
\varepsilon^2 = \mathbb E \left[ \ell\left(\hat \Sigma, 
\mathbb E[\hat \Sigma]\right) \right]
\end{equation}

Under the MSE loss function $\mathcal L_e$ (\autoref{eq:MSE}), the decomposition is the simple sum $r = b^2 + \varepsilon^2$. 
Under other loss functions, $r$ is an increasing function of both $b^2$ and $\varepsilon^2$ although not generally a simple sum.  
\TODO{As far as I know, the terms `bias' and `variance' are only applicable under a quadratic loss function  like $\mathcal L_e$.  Under other loss functions, we may need to use approximation/estimation error.}

@@

\subsection{Regularization}

@@

A number of regularization schemes have been developed, which differ by their target estimates and mixing functions.    Curiously, \emph{some} improvement can be produced with an arbitrary target estimate as long as its variance is lower than that of $\hat\Sigma_0$.  The perplexing phenomenon that estimates can be improved by a bias toward an arbitrary less variable target is known as \emph{Stein's paradox} \citep{Efron:1977}.   \TODO{This point can be untuitive.  For example, \cite{Varoquaux:2012} misrepresents the effect of regularization by stating that regularization removes the upward bias of correlations in the sample covariance matrix. In reality, regularization should start with an unbiased estimator and bias it toward a low-variance target so that the estimation risk is reduced.} However, if a family of target estimates can capture the underlying regularities in the covariance structure of a specific system with a small number of parameters, biasing the estimate toward these targets ought to produce a greater reduction in estimator risk.

>>>>

\subsection*{Basic concepts}
Let $x(t) \in \mathcal X, t=1\ldots,n$ denote a sample of consecutive observations of population activity of $p$ neurons in time bins $t$.  
The activity of each neuron is denoted by the real-valued firing rate, thus  $\mathcal X = \mathbb R^{p\times 1}$.  
For example, in our applicaiton, the firing rate is estimated from somatic calcium fluorescence signals with the mean stimulus response subtracted, making it possible for elements of $x(t)$ to take on negative and non-integer values.

@@

But the usual estimator of the true covariance matrix $\Sigma$ is  the sample covariance matrix:
\begin{equation}
\hat\Sigma_0 =  \frac 1 {n-c} \sum\limits_{t=1}^n (x(t) -\hat\mu) (x(t) - \hat\mu)^\T
\quad\mbox{where}\;
\hat\mu = \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $c$ is the sum of temporal correlations \TODO{there must be a better description of this}. For independent observations, $c=1$;  $c>1$ when nearby samples are correlated; the value of $c$ can be estimated from the data.
\TODO{Under the Gaussian loss, $\mathcal L_g$ (\ref{eq:GaussLoss}) the scaling is irrelevant, so we will not spend any time talking about the scaling.  But it would be important if we used $\mathcal L_e$ (\ref{eq:MSE}).}

@@

%Nonlinear functions of an unbiased estimate can be biased: $\mathbb E\left[\hat\Sigma_0\right]=\Sigma \centernot\implies  E\left[\varphi(\hat\Sigma_0)\right]=\varphi(\Sigma)$.  For example, if $u_{\max}(S)$ is the largest eigevalue of square matrix $S$, then $\mathbb E\left[u_{\max}(\hat\Sigma_0)\right] > u_{\max}(\Sigma)$. Some covariance matrix estimators are designed to correct for the eigenspectrum bias at the cost  of adding bias to covariance coefficients \cite{Ledoit:2004}.

\subsection*{Loss and risk}
The optimization of a covariance matrix estimate $\hat\Sigma$ is performed with respect to a \emph{loss function} $\mathcal L(\hat\Sigma,\Sigma)$, which expresses the discrepancy between $\hat\Sigma$ and $\Sigma$ and attains its minimum when $\hat\Sigma=\Sigma$.  
Then \emph{excess loss}  
\begin{equation}
\ell(\hat\Sigma,\Sigma) = \mathcal L(\hat\Sigma,\Sigma)-\mathcal L(\Sigma,\Sigma)
\end{equation}
assumes zero at its minimum.

@@

The Gaussian loss function $\mathcal L_g$  arises from the theory of multivariate normal distributions. When observations are identically and independently distributed according to a multivariate normal distribution with zero means, the log likelihood of the covariance matrix $\Sigma$ with $\hat\Sigma = \frac 1 n \sum\limits_{t=1}^n x(t) x(t)^\T$ is  
\begin{equation}
L\left(\Sigma \mid \hat\Sigma\right) = -\frac n 2 \ln(2\pi) - \frac n 2 \ln \det \Sigma - \frac n 2 \Tr(\Sigma^{-1} \hat \Sigma)
\end{equation}
Then $\mathcal L_g$ is constructed by rescaling $L\left(\Sigma \mid \hat\Sigma\right)$ and dropping the constant term:
\begin{equation}\label{eq:GaussLoss}
\mathcal L_g(\hat\Sigma,\Sigma) 
=  -\frac 2 {pn} L\left(\Sigma \mid \hat\Sigma \right) - \frac 1 p \ln(2\pi) 
\equiv  \frac 1 p\left(\ln \det \hat \Sigma + \Tr(\hat \Sigma^{-1}) \right) 
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat\Sigma,\Sigma) = \mathcal L_g(\hat\Sigma,\Sigma) - \mathcal L_g(\Sigma,\Sigma)  
= \frac 1 p \left(-\ln \det (\hat \Sigma^{-1} \Sigma) + \Tr(\hat \Sigma^{-1}\Sigma)\right) - 1
\end{equation}
is known as \emph{entropy loss} \cite{James:1961}. \TODO{The order of the operands is reversed. Check that it's okay.}

@@

\subsection*{Cross validation}
In practice, the true covariance matrix $\Sigma$ is not accessible and the estimator risk $r$ must be estimated from data. This requires a separate \emph{validation} empirical distribution $\hat F_m^\prime$ of $m$ observations sampled from $F$ independently of the \emph{training} distribution $\hat F_n$. 
Then let  $\hat \Sigma_0^\prime$ be the sample covariance obtained from $\hat F_m^\prime$.

@@

\subsection*{Bias/variance decomposition}
The estimator risk (\ref{eq:risk}) can be decomposed into \emph{approximation error} or \emph{bias}   
\begin{equation}
b^2 = \ell \left( \mathbb E[\hat \Sigma],\Sigma\right)
\end{equation}
and \emph{estimation error} or \emph{variance}
\begin{equation}
\varepsilon^2 = \mathbb E \left[ \ell\left(\hat \Sigma, 
\mathbb E[\hat \Sigma]\right) \right]
\end{equation}

Under the MSE loss function $\mathcal L_e$ (\ref{eq:MSE}), the decomposition is the simple sum $r = b^2 + \varepsilon^2$. 
Under other loss functions, $r$ is an increasing function of both $b^2$ and $\varepsilon^2$ although not generally a simple sum.  
\TODO{As far as I know, the terms `bias' and `variance' are only applicable under a quadratic loss function  like $\mathcal L_e$.  Under other loss functions, we may need to use approximation/estimation error.}

@@

\subsection*{Regularization}

@@

A number of regularization schemes have been developed, which differ by their target estimates and mixing functions.    Curiously, \emph{some} improvement can be produced with an arbitrary target estimate as long as its variance is lower than that of $\hat\Sigma_0$.  The perplexing phenomenon that estimates can be improved by a bias toward an arbitrary less variable target is known as \emph{Stein's paradox} \cite{Efron:1977}.   \TODO{This point can be untuitive.  For example, \cite{Varoquaux:2012} misrepresents the effect of regularization by stating that regularization removes the upward bias of correlations in the sample covariance matrix. In reality, regularization should start with an unbiased estimator and bias it toward a low-variance target so that the estimation risk is reduced.} However, if a family of target estimates can capture the underlying regularities in the covariance structure of a specific system with a small number of parameters, biasing the estimate toward these targets ought to produce a greater reduction in estimator risk.
